Idiosyncratic, Ordinal Tax To Sustain a Digital Economy

### 2019-08-04 writing/short note on digital tax/DigitalTax.tex
minor changes in short note.

Recently, the French Senate approved a law that imposes a $3\%$ tax on revenue generated from digital services by companies above a certain size. 
While there is a lot of debate about political consequences of this action, it is actually interesting to reverse the question and instead ask what are the long-term implications of an economy with no such digital tax. 
More generally, we can see a digital service as a special case of products with low or zero cost of transportation. 
With basic economic models we show that an economy with no transportation costs is prone to monopolisation in that minuscule, random differences in quality are rewarded disproportionally. 
We then propose a distance-based tax to counter-balance the tendencies of random centralisation.  
Compared to an obvious tax that scales with physical (cardinal) distance, we show that a ranked (ordinal) distance tax leverages the benefits of digitalisation while keeping the economy stable.

@@

If there is no cost of transportation, the difference in profit is zero as predicted by symmetry (both firms are exactly equal).
However, this equality is highly unstable. 
To see this, let us now consider the case where firm $1$ is marginally more competitive ($c_1 = 99, c_2 =100$). 
As long as there is a cost of transportation $(\lambda=0.1)$, the difference in profit merely increases from $5\%$ to $7\%$.
But in absence of such a cardinal tax ($\lambda=0$), the difference in revenue amounts to a whopping $33\%$!
This result is particularly bothersome as this excess profit can be used to further decrease costs of production.
Either through genuine innovation, or temporary undercutting of prices, the difference between $c_1$ and $c_2$ is further enlarged and the second firm is run out of business quickly. 
\footnote{
This claim can be formalised by considering multi-stage games.
} 
In general, the outpace of competition is natural and in many cases even desired, as efficiency is to be rewarded. 
However, we stress that a minor, essentially random difference of $1\%$ production cost disproportionally benefited one firm. 
We may rationalise this by going back to the example of self-contained villages from the introduction: 
Once there is no more cost of transportation, the economic agent has little incentive to favour local bread, if bread from another village is just marginally cheaper. 
Monopolisation initiated by minor, random initial differences in quality is thus to be expected.

These ever decreasing (or non-existing) cost of transportation can be counter-balanced by introducing a tax that scales with distance (e.g. a carbon tax). 
An immediate side-effect of such market interference is less consumption and hence decreased revenue for both firms. (second vs. third bar on right axis in Figure \ref{fig:results}). 
While decentralised, locally self-contained economies may benefit consumers on a grand scheme, it certainly punishes remote buyers and is susceptible to potentially small changes in location. 
Therefore, we propose to introduce an \textit{ordinal tax} instead. 
For each buyer, the distance to the different firms is enumerated in increasing order. 
A buyer has a distance of $D=0$ from the physically (cardinally) closest firm, a distance $D=1$ from the second closest, and so forth. 
The advantages of this tax are manifold. 
First, it resolves the issue of remote buyers being worse off. 
Second, it is not susceptible to (small) changes in buyer or firm locations, and hence does not lead to physical centralisation (profits are exactly equal if $c_1=c_2$). 
Third, the overall level of revenue is elevated compared to a cardinal tax, as now only remote purchases are taxed (right axis in Figure \ref{fig:results}). 
Fourth, marginal differences in quality remain unnoticed, as the difference in production cost does not outweigh taxation costs. 
Fifth, major difference in quality (true competitive advantage), overcomes the tax and rewards the more competitive firm 
(cf. in Figure \ref{fig:results} the $50$ times higher revenue in case of a $20\%$ difference in production cost).

>>>>

Recently, the French Senate approved a law that imposes a $3\%$ tax on revenue generated from digital services by companies above a certain size. 
While there is a lot of debate about political consequences of this action, it is actually interesting to reverse the question and instead ask what are the long-term implications of an economy with no such digital tax. 
More generally, we can see a digital service as a special case of products with low or zero cost of transportation. 
With basic economic models we show that an economy with no transportation costs is prone to monopolisation in that minuscule, random differences in quality are rewarded disproportionally. 
We then propose a distance-based tax to counter-balance the tendencies of random centralisation.  
Compared to an obvious tax that scales with physical (cardinal) distance, we show that a ranked (ordinal) distance tax leverages the benefits of digitalization while keeping the economy stable.

@@

If there is no cost of transportation, the difference in profit is zero as predicted by symmetry (both firms are exactly equal).
However, this equality is highly unstable. 
To see this, let us now consider the case where firm $1$ is marginally more competitive ($c_1 = 99, c_2 =100$). 
As long as there is a cost of transportation $(\lambda=0.1)$, the difference in profit merely increases from $5\%$ to $7\%$.
But in absence of such a cardinal tax ($\lambda=0$), the difference in profit amounts to a whopping $33\%$!
This result is particularly bothersome as this excess profit can be used to further decrease costs of production.
Either through genuine innovation, or temporary undercutting of prices, the difference between $c_1$ and $c_2$ is further enlarged and the second firm is eventually run out of business. 
\footnote{
This claim can be formalised by considering multi-stage games.
} 
In general, the outpace of competition is natural and in many cases even desired, as efficiency is to be rewarded. 
However, we stress that a minor, essentially random difference of $1\%$ production cost disproportionally benefited one firm. 
We may rationalise this by going back to the example of self-contained villages from the introduction: 
Once there is no more cost of transportation, the economic agent has little incentive to favour local bread, if bread from another village is just marginally cheaper. 
Monopolisation initiated by minor, random initial differences in quality is thus to be expected.

These ever decreasing (or non-existing) cost of transportation can be counter-balanced by introducing a tax that scales with distance (e.g. a carbon tax). 
An immediate side-effect of such market interference is less consumption and hence decreased revenue for both firms. (second vs. third bar on right axis in Figure \ref{fig:results}). 
While decentralised, locally self-contained economies may benefit consumers on a grand scheme, it certainly punishes remote buyers and is susceptible to potentially small changes in location. 
Therefore, we propose to introduce an \textit{ordinal tax} instead. 
For each buyer, the distance to the different firms is enumerated in increasing order. 
A buyer has a distance of $D=0$ from the physically (cardinally) closest firm, a distance $D=1$ from the second closest, and so forth. 
The advantages of this tax are manifold. 
First, it resolves the issue of remote buyers being worse off. 
Second, it is not susceptible to (small) changes in buyer or firm locations, and hence does not lead to physical centralisation (profits are exactly equal if $c_1=c_2$). 
Third, the overall level of revenue is elevated compared to a cardinal tax, as now only remote purchases $(D > 0)$ are taxed (right axis in Figure \ref{fig:results}). 
Fourth, marginal differences in quality remain unnoticed, as the difference in production cost does not outweigh taxation costs. 
Fifth, major difference in quality (true competitive advantage), overcomes the tax and rewards the more competitive firm 
(cf. in Figure \ref{fig:results} the $50$ times higher revenue in case of a $20\%$ difference in production cost).
